Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-11 TOMS_Paper/assembly.tex
Add some entries in autotuner table, still incomplete
Minor tweaks

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become incredibly popular over the last years. In these frameworks, the weak variational form of a given problem is expressed at high-level by means of a domain-specific language. Such a mathematical specification is suitably manipulated and then passed as input to a form compiler, whose goal is to generate a representation of local assembly operations. These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution. Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more expensive, covering a significant fraction of overall computation run-time.

Achieving high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Also, traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. This has led to development of a number of higher-level approaches to optimize local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of the problem. In~\cite{Luporini}, we have made one step forward by showing that different problems, on different platforms, require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu}.

@@

\newcommand{\autotuneroutput}{
\tiny
\begin{tabulary}{1.0\textwidth}{c|c|ccc|ccccc}
\cline{3-10}
\multicolumn{2}{c}{} & \multicolumn{3}{c|}{\texttt{expression rewriting}} & \multicolumn{5}{c}{\texttt{code specialization}} \\
\cline{1-10}
problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & BLAS \\
\cline{1-10}
\texttt{helmholtz} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{mass} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{elasticity} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{poisson} & 48 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{mixed poisson} & 40 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{hyperelasticity} & 32 & 1 & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\cline{1-10}
\end{tabulary}
}

>>>>

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS~\cite{Fenics} and Firedrake~\cite{firedrake-code} projects, which have become incredibly popular over the last years. In these frameworks, the weak variational form of a given problem is expressed at high-level by means of a domain-specific language. Such a mathematical specification is suitably manipulated and then passed as input to a form compiler, whose goal is to generate a representation of local assembly operations. These operations numerically evaluate problem-specific integrals in order to compute so called local matrices and vectors, which represent the contributions from each element in the discretized domain to the equation solution. Local assembly code must be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, and polynomial order of the chosen function spaces, the resulting assembly kernels become more and more computationally expensive, covering a significant fraction of the overall computation run-time.

Achieving high performance implementations is, however, non-trivial. The complexity of mathematical expressions involved in the numerical integration, which varies from problem to problem, and the small size of the loop nest in which such integral is computed obstruct the optimization process. Also, traditional vendor compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent assembly expressions. This has led to development of a number of higher-level approaches to optimize local assembly kernels.  In~\cite{quadrature1}, it is shown how automated code generation can be leveraged to introduce domain-specific optimizations, which a user cannot be expected to write ``by hand''. \cite{Kirby-FEM-opt} and~\cite{Francis} have studied, instead, different optimization techniques based on a mathematical reformulation of the problem. In~\cite{Luporini}, we have made one step forward by showing that different problems, on different platforms, require distinct set of transformations if close-to-peak performance needs to be obtained, and that low-level, domain-aware code transformations are essential to maximize instruction-level parallelism and register locality. The problem of optimizing local assembly routines has been tackled recently also for GPUs, for instance in~\cite{petsc-integration-gpu},~\cite{Klockner}, and~\cite{Bana}.

@@

\newcommand{\autotuneroutput}{
\tiny
\begin{tabulary}{1.0\textwidth}{c|c|ccc|ccccc}
\cline{3-10}
\multicolumn{2}{c}{} & \multicolumn{3}{c|}{\texttt{expression rewriting}} & \multicolumn{5}{c}{\texttt{code specialization}} \\
\cline{1-10}
problem & \specialcell{number of\\variants} & \specialcell{rewrite\\strategy} & \specialcell{zero-valued columns\\avoidance} & \specialcell{precompute\\integration terms} & \specialcell{padding\\data alignment} & split & \specialcell{vector-register\\tiling} & unroll & BLAS \\
\cline{1-10}
\texttt{helmholtz} & 64 & aggressive & 0 & 0 & all & 8 & 17 & 15 & 0 \\
\texttt{mass} & 64 & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{elasticity} & 64 & aggressive & 39 & 0 & all & 0 & 11 & 0 & 2 \\
\texttt{poisson} & 64 & aggressive & 50 & 0 & all & X & X & 1 & 0 \\
\texttt{mixed poisson} & X & base & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\texttt{hyperelasticity} & 32 & aggressive & 2 & 3 & 1 & 2 & 3 & 4 & 5 \\
\cline{1-10}
\end{tabulary}
}
